---
title: "Homework Assignment 6"
author: "Santiago Garcia"
date: "May 22, 2023" 
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

## Question 1 (5.16)

a. 

- **Alternative:** Has the calorie intake changed from 1100 since they started putting calorie counts in each item in the restaurant $H_a: p \neq 0.5$

- **Null:** has the calorie intake stayed the same every since they started putting calorie counts in each item in the restaurants  $H_o: p =  0.5$

b.

- **Alternative:** On average do more or less than 70& of Wisconsin residents drink alcohol $H_a: p \neq 0.7$

- **Null:** Do 70% of Wisconsin residents drink alcohol $H_o: p = 0.7$

## Question 2 (5.25)

a. Does the anti depressants effective at help Diana with her Symptoms?

b. Type 1 error would be that the drug is helping her while in fact it isn't 

b. Type 2 error would be that stating that the drug does not help her, while in fact it is helping her

## Question 3 (5.30)

a. True, our 95% percent confidence interval is also within our 99%, if we are 99% sure then that means we are including that 95% percent in our statement 

b. False, the probability of causing a type 1 error is alpha thus, they have a direct relationship as alhpa increases the more likely we are to commit a type error, not the way around

c. False, If we fail to reject $H_0$ this means that the true population proportion is 0.5

d. True, 

## Question 4 (6.4)

a. 

$pn >= 10 \\0.25 * 12 >= 10 \\ 3 >= 10$

b. True

Checks for normal distribution:

$np >= 10 \\ 0.25 * 40 >= 10 \\ 10 >=10 \\ True$
$n(1- p) >= 10 \\ = 40(1-0.25 >= 10) \\ = 30 >= 10 \\ True$

This sample size would probably gives us a normal distribution for sample proportion

c. False

$\hat{p} \pm cv * SE$
$0.25 \pm 1.96 * 0.12$
(0.13, 0.37)

the value of 20% of is within our confidence interval

d. False

$\hat{p} \pm cv * SE$
$0.25 \pm 0.0693$
(0.1807, 0.2193)

the value of 20% is within our confidence interval 

e. False

$SE = \frac{\sqrt{\hat{p}(1 - \hat{p})}}{\sqrt{n}}$

if you triple the sample size it will reduce the standard error by $\frac{1}{\sqrt{3}}$, in order to reduce standard error by a third you would need to multiple sample size by 9

## Question 5 (6.10)

a. 61% is a statistic since they are getting the results froma survey 

b. 

$\hat{p} \pm cv * SE$
$0.61 \pm 0.0123$

(0.5977, 0.6223) We are 95% certain that between 59.77% and 62.23% of U.S residents think the use of marijuana should be made legal

c. 

$np >= 10$
$962.58 >= 10$ True

$n(1-p) >= 10$
$615.42 >= 10$ True 

The data pass the normal distribution checks, thus is should follow a normal distribution or nearly normal


d. 

$H_o: p <= 0.5$
$H_a: p > 0.5$

$Z = \frac{\hat{p} - p_o}{SE}$
$Z = 8.7393$

p(Z > z)
1 - p(z < 8.7393)

```{r}
1 - pnorm(9.73923)
```

p <= a

0 <= 0.5 (True)

We reject $H_o$ at a 5% significance, there is enough evidence to believe that the majority of Americans believe that marijuana should be made legal the United States. 


## Question 6 (6.12)

a.

$H_o: p >= 0.5$
$H_a: p < 0.5$

$Z = \frac{\hat{p} - p_o}{SE}$
$Z = -0.72$

p(Z < z)

p(z < -0.7277)

```{r}
pnorm(-0.7277)
```
$P_{val} > \alpha$
$ 0.233 > 0.05$

We fail to reject $H_o$, thus there is not enough evidence to prove that a minority of Americans that did not go to college was because they could not afford it 

b. I would expect the value of 0.5 to be in our confidence interval because we fail to reject $H_o$ thus this means that our proportion is going to contain values of 0.5 or more 

## Question 7 (6.14)

a. 

$\hat{p} \pm cv * SE$
$0.48 \pm 1.64 * 0.0264$

(0.4536, 0.5233), We are 90% confident that between 45.36% and 52.33% of Americans that did not go to college was because they could not afford it. 

b. 

$n = \frac{CV^2 X \hat{p}(1-\hat{p})}{E^2}$

$= \frac{1.64^2 X 0.48(1-0.48)}{0.015^2}$

= 2983.66293

We need a sample size of **2984** to have a margin of error of 1.5%

## Question 8 (6.16)

a. 

$n = \frac{CV^2 X \hat{p}(1-\hat{p})}{E^2}$
$= \frac{1.96^2 X 0.61(1-0.61)}{0.02^2}$

= 2284.7916

We need a to survey **2285** in order to get a margin error of 2% 

