---
title: "Homework Assignment 5"
author: "Santiago Garcia"
date: "May 15, 2023" 
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
---

```{r}
library("tidyverse")
library("statsr")
library("broom")
lakeData <- read.csv("Lake.csv")
```


## Question 1:

a. The dependent variable would be Mercury because these are things that are measured in the lake, and pH is calculated due to the Mercury of the lake 

b. 

```{r warning=FALSE}
ggplot(lakeData, aes(x = pH, y = Mercury)) + geom_point()

ggplot(lakeData, aes(x = Mercury, y = pH)) + geom_boxplot()
```
There is a linear moderate negative relationship between Mercury and pH, and there are not outliers 

c. 

```{r}
correlation <- cor(x = lakeData$pH, y = lakeData$Mercury)

determination <- correlation ^ 2

```

This implies that pH levels contributes 37% toward your Mercury levels in the water.

d. 

```{r}
model <- lm(Mercury ~ pH, lakeData)
model2 <- lm(pH ~ Mercury, lakeData)
summary(model)
```
$\hat{Mercury} = 1.4878 - 0.1496pH$

**slope:** for a unite increase in pH, it will causes a unit decrease of Mercury by the value 0.14896

**intercept:** if pH is 0, then 1.4878 is the mean value of Mercury 



e. 

```{r}
resid <- residuals(model)
hist(resid)

aug2 <- augment(model)
ggplot(data = aug2) +
    aes(x = .fitted, y = .resid) +
    geom_point() + 
    geom_hline(yintercept = 0, color = "blue") +
  geom_smooth() + 
    ggtitle("Residuals vs. Fitted") +
    xlab("Fitted Values") +
    ylab("Residuals")

```
Linearity: scatterplot, and r coefficient show that the relation is linear

Normality: the model shows almost nearly normal residuals 

Constant Variability: the graph shows that there is no constant variability


## Question 2:

a. 
```{r}
pnorm(81, 79, 4) - pnorm(77, 79, 4)
```

p(77 <= x <= 81) = p(x <= 81) - p(x <= 77)

= 0.3829

b.  

since this is a normal distribution, no matter how big our sample size is, we are still going to get the same distribution of heights, our mean staying the same and also standard deviation

c. 

1.5 * SD = 6

$SD \pm 6$ = 73, 85

p(73 <= x <= 85)

```{r}

pnorm(85, 79, 4) - pnorm(73, 79, 4)
```
```{r}
#85% cutoff
qnorm(0.85, 79, 4)

#15% cutoff
qnorm(0.15, 79, 4)

```

the middle 70% cutoffs would be in between 15% and 85%, and thus are 

15%: 74.8542 inches and 85%: 83.1457 inches, 85% - 15% would be the 70%


## Question 3: 

a. the parameter would be American who used wireless means for an online access 

$\hat{p} = \frac{1226}{2235}$
$= 0.5485$

the statistic to measure the uncertainty of the point estimate is called the margin of error 

b. 

$Z^* = Z\frac{alpha}{2}$

$= Z0.025$

```{r}
qnorm(0.025, 0, 1)
```
= -1.96

$E = Z^*\sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}$
$E = Z^*\sqrt{\frac{{0.5485}(1 - {0.5485})}{2235}}$
= 0.0206

$\hat{p} \pm Z^*SE$

$0.5485 \pm0.0206$

= (0.5279, 0.5691)

We are 95% confident for all american adults who is wireless means is between 52.79% and 56.91% 

c. Yes, he should be surprised because his value is not within our CI

d. 

$n = \frac{z^x * \hat{p}(1-\hat{p})}{e^2}$
$n = \frac{-1.96^2 * 0.5485(1-0.5485)}{0.03^2}$
$n = 1058$ people


